---
title: "1 · Data & preparation"
subtitle: "One canonical preparation, every decision in the open"
---

Every chapter in this suite asks a different question of the same 9,841 addresses —
which makes the preparation of those addresses the one place where quiet
inconsistency could poison everything downstream. A correlation tested on one
cleaning of the data and a regression fitted on a subtly different one are not
results about the same population. So preparation is treated here as a first-class
analysis step: the raw file is loaded once, every cleaning decision is made once —
stated and justified where a reader can disagree with it — and the result is
exported as the single starting point for the statistical chapters. The Python
exploration deliberately begins from the same raw file instead, walking the data by
hand; that is where these decisions are *earned* rather than asserted.

## Source

The [Kaggle Ethereum Fraud Detection dataset](https://www.kaggle.com/datasets/vagifa/ethereum-frauddetection-dataset):
one row per Ethereum address, aggregating its transaction history into activity,
value, timing and ERC20-token features, with `FLAG = 1` marking addresses known to
have engaged in fraudulent activity.

```{r load}
library(dplyr)

trxs_raw <- read.csv("data/transaction_dataset.csv")
dim(trxs_raw)
```

## Canonical column names

`read.csv` mangles the original headers (spaces and punctuation become dots, a
leading space becomes an `X.` prefix). Everything is normalised to snake_case once,
and the one unwieldy name the analyses use constantly is shortened.

```{r names}
nm <- tolower(names(trxs_raw))
nm <- gsub("\\.+", "_", nm)        # runs of dots -> single underscore
nm <- gsub("^x_|^_|_$", "", nm)    # strip artefact prefixes / stray underscores
names(trxs_raw) <- nm

trxs <- trxs_raw %>%
  rename(total_transactions = total_transactions_including_tnx_to_create_contract)
```

## Cleaning decisions

Four decisions shape the analysis sample — each written down precisely so a reader
can push back on it.

**Fraud label.** Kept both as the numeric `flag` (for modelling) and as a labelled
factor `flag_f` (for tables and plots). The class balance is retained as-is — the
imbalance is a property of the phenomenon, and chapters that are sensitive to it
(the logistic regression in particular) address it explicitly rather than resample
it away.

**ERC20 token types.** The two categorical token columns are polluted with
placeholder values — `""`, `"0"`, and whitespace variants — which all mean the same
thing: *no dominant ERC20 token*. They are unified into an explicit `"None"`
category rather than dropped, because "no ERC20 activity" is itself informative.

**Missing ERC20 aggregates.** 829 addresses have `NA` across the ERC20 aggregate
columns. They are **kept** here; chapters drop them only where a model requires
those variables, and say so. This quirk turns out to be more than a quirk —
Chapter 5 returns to it with a finding worth the wait.

**Negative balances.** 978 addresses show a negative `total_ether_balance`, an
artefact of how the aggregate was computed at extraction time rather than a
possible state of an account. They too are **kept** here and handled where they
matter — the linear-regression chapter, which excludes them with justification.

```{r clean}
clean_token <- function(x) {
  x <- trimws(as.character(x))
  ifelse(is.na(x) | x %in% c("", "0", "None"), "None", x)
}

trxs <- trxs %>%
  mutate(
    flag_f = factor(flag, levels = c(0, 1),
                    labels = c("Legitimate", "Fraudulent")),
    erc20_most_sent_token_type = clean_token(erc20_most_sent_token_type),
    erc20_most_rec_token_type  = clean_token(erc20_most_rec_token_type)
  )
```

## The dataset at a glance

```{r quality}
quality <- tibble(
  check = c("Addresses (rows)", "Features (columns)",
            "Flagged fraudulent", "Fraud share",
            "Missing ERC20 aggregates", "Negative total_ether_balance",
            "Token type = 'None' (most received)"),
  value = c(nrow(trxs), ncol(trxs),
            sum(trxs$flag), sprintf("%.1f%%", 100 * mean(trxs$flag)),
            sum(is.na(trxs$total_erc20_tnxs)),
            sum(trxs$total_ether_balance < 0, na.rm = TRUE),
            sum(trxs$erc20_most_rec_token_type == "None"))
)
knitr::kable(quality, caption = "Dataset at a glance, after canonical preparation.")
```

## Key variables used downstream

Fifty features arrive with the data; the analyses lean on a focused subset, mapped
here so every later chapter can name a variable without re-introducing it.

```{r glossary, echo=FALSE}
glossary <- tribble(
  ~variable, ~concept, ~type, ~used_in,
  "flag / flag_f", "Known-fraud label", "Binary / factor", "Ch. 3, 4, 5",
  "total_ether_sent", "Total Ether sent (per address)", "Continuous", "Ch. 3, 4",
  "total_transactions", "Total transactions incl. contract creation", "Continuous", "Ch. 3, 4, 5",
  "avg_val_received / avg_val_sent", "Average Ether value received / sent", "Continuous", "Ch. 3",
  "avg_min_between_sent_tnx", "Average minutes between sent transactions", "Continuous", "Ch. 3, 5",
  "total_ether_balance", "Ether balance after all transactions", "Continuous", "Ch. 4",
  "total_erc20_tnxs", "Total ERC20 token transactions", "Continuous", "Ch. 5",
  "erc20_most_sent_token_type", "Most frequently sent ERC20 token", "Nominal", "Ch. 3",
  "erc20_most_rec_token_type", "Most frequently received ERC20 token", "Nominal", "Ch. 3, 5"
)
knitr::kable(glossary, caption = "Glossary of variables used in the analysis chapters.")
```

## Export the canonical dataset

```{r export}
saveRDS(trxs, "data/eth_clean.rds")
write.csv(trxs, "data/eth_clean.csv", row.names = FALSE)
```

The statistical chapters start from `readRDS("data/eth_clean.rds")` and inherit
every decision above without repeating it. The Python exploration chapter
intentionally does not: it works from the same raw file and performs its own
cleaning in full view, because watching the data get cleaned — the missingness,
the placeholder values, the tails — is where the decisions codified here stop
looking arbitrary. Two tracks, one raw source, consistent pictures.
