---
title: "3 · Hypothesis testing"
subtitle: "Five research questions on fraud-relevant transaction behaviour"
---

Chapter 2 ended with patterns an analyst can *see*: heavy-tailed value
distributions, timing differences between the flagged and legitimate populations,
token activity that looks fraud-linked, and value features that move together. Eyes
are not evidence, though — apparent patterns in a 9,841-address snapshot can be
noise, scale artefacts, or both. This chapter subjects five of those observations
to formal statistical testing.

The five research questions fall into three natural groups. Two are about **how
value and activity relate** — whether busier addresses send more Ether (RQ1) and
whether addresses operate at a characteristic value scale (RQ2). One is about
**behaviour over time** — whether fraudulent and legitimate addresses transact on
different rhythms (RQ3). And two are about **token ecosystems** — whether Ether
outflow differs across dominant sent-token types (RQ4) and whether the dominant
received token is associated with the fraud flag itself (RQ5). Each question gets
the same discipline: hypotheses stated up front, assumptions checked against an
explicit criterion, the appropriate test run — and a second estimator that does not
share the first one's assumptions, so that no conclusion in this chapter hinges on
a distributional judgement call.

## Ground rules for every test

Two decisions apply throughout. First, the **working scale**: every value and
volume variable here is heavily right-tailed (Chapter 2 made that vivid), so tests
that assume anything about distribution shape operate on `log10(x + 1)` scales,
while rank-based checks use the raw values they are indifferent to. Second, the
**normality criterion**, stated explicitly rather than judged by eye: a transformed
variable is treated as approximately normal for large-sample purposes when its
standardised skew and kurtosis stay within ±3.29 and no more than 1% of
standardised scores exceed |3.29|. The helper below implements exactly that
assessment, and a small plotting helper shows each distribution against its normal
reference.

```{r setup}
library(dplyr)
library(ggplot2)
library(car)

trxs <- readRDS("data/eth_clean.rds")

# ---- Normality assessment helper -------------------------------------------
# Standardised skew & kurtosis plus the share of |z| beyond 1.96 / 3.29 —
# the large-sample criteria described above, made explicit and reusable.
normality_summary <- function(x, label) {
  x <- x[is.finite(x)]
  n <- length(x); m <- mean(x); s <- sd(x)
  z <- (x - m) / s
  m2 <- mean((x - m)^2); m3 <- mean((x - m)^3); m4 <- mean((x - m)^4)
  skew <- m3 / m2^1.5
  kurt <- m4 / m2^2 - 3
  se_skew <- sqrt(6 * n * (n - 1) / ((n - 2) * (n + 1) * (n + 3)))
  se_kurt <- 2 * se_skew * sqrt((n^2 - 1) / ((n - 3) * (n + 5)))
  tibble(variable = label, n = n, mean = m, sd = s,
         skew = skew, skew_std = skew / se_skew,
         kurtosis = kurt, kurt_std = kurt / se_kurt,
         `%|z|>1.96` = 100 * mean(abs(z) > 1.96),
         `%|z|>3.29` = 100 * mean(abs(z) > 3.29))
}

# ---- Distribution plot helper ----------------------------------------------
dist_plot <- function(data, var, label) {
  ggplot(data, aes(x = .data[[var]])) +
    geom_histogram(aes(y = after_stat(density)), bins = 40,
                   colour = "black", fill = "grey80") +
    stat_function(fun = dnorm, colour = "red",
                  args = list(mean = mean(data[[var]], na.rm = TRUE),
                              sd = sd(data[[var]], na.rm = TRUE))) +
    labs(x = label, y = "Density") +
    theme_minimal()
}

# Log-transformed working copies for the shape-sensitive tests.
trxs <- trxs %>%
  mutate(
    total_ether_sent_log    = log10(total_ether_sent + 1),
    total_transactions_log  = log10(total_transactions + 1),
    avg_val_received_log    = log10(avg_val_received + 1),
    avg_val_sent_log        = log10(avg_val_sent + 1),
    avg_min_sent_log        = log10(avg_min_between_sent_tnx + 1)
  )
```

## RQ1 — Do addresses that transact more also send more Ether?

Chapter 2 showed volume and frequency both exploding across orders of magnitude;
the first question is whether they explode *together*. If they largely do, one is
redundant given the other; if they only loosely do, each carries its own signal.

- **H0:** there is *no* relationship between total Ether sent and total number of
  transactions per address.
- **Ha:** there *is* a relationship between total Ether sent and total number of
  transactions per address.

```{r rq1-normality}
bind_rows(
  normality_summary(trxs$total_ether_sent_log,   "log10(total_ether_sent + 1)"),
  normality_summary(trxs$total_transactions_log, "log10(total_transactions + 1)")
) %>% knitr::kable(digits = 2,
                   caption = "Normality assessment of the transformed variables against the ±3.29 criterion.")
```

```{r rq1-plots}
#| layout-ncol: 2
#| fig-cap:
#|   - "Total Ether sent after log transformation, with its normal reference curve."
#|   - "Total transactions after log transformation, with its normal reference curve."
dist_plot(trxs, "total_ether_sent_log",   "log10(total Ether sent + 1)")
dist_plot(trxs, "total_transactions_log", "log10(total transactions + 1)")
```

```{r rq1-tests}
rq1_pearson <- cor.test(trxs$total_ether_sent_log, trxs$total_transactions_log,
                        method = "pearson")
rq1_kendall <- cor.test(trxs$total_ether_sent, trxs$total_transactions,
                        method = "kendall")
rq1_pearson
```

```{r rq1-scatter}
#| fig-cap: "Each point is an address. The association is clearly positive — and clearly loose: at any activity level, outflow still spans several orders of magnitude."
ggplot(trxs, aes(total_transactions_log, total_ether_sent_log)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_smooth(method = "lm", colour = "red") +
  labs(title = "Outflow rises with activity — but loosely",
       x = "log10(total transactions + 1)", y = "log10(total Ether sent + 1)") +
  theme_minimal()
```

**Finding.** A statistically significant positive association:
Pearson *r* = `r round(rq1_pearson$estimate, 3)`
(*p* `r format.pval(rq1_pearson$p.value, eps = .001)`) on the transformed scales,
with Kendall's τ = `r round(rq1_kendall$estimate, 3)` on the raw ranks confirming
direction and significance. The association is real but **weak-to-moderate** —
transaction *count* alone is a poor proxy for Ether *volume*. That looseness is
itself informative: volume- and frequency-based signals carry partially independent
information about an address, a fact the regression chapters will lean on.

## RQ2 — Do high-value receivers also send high values?

The second value question is about *scale of operation*: does an address that
receives large average amounts also send large average amounts — retail behaviour
staying retail, whale behaviour staying whale?

- **H0:** there is no positive correlation between the average Ether value received
  and the average Ether value sent.
- **Ha:** there is a positive correlation between the average Ether value received
  and the average Ether value sent.

```{r rq2-normality}
bind_rows(
  normality_summary(trxs$avg_val_received_log, "log10(avg_val_received + 1)"),
  normality_summary(trxs$avg_val_sent_log,     "log10(avg_val_sent + 1)")
) %>% knitr::kable(digits = 2,
                   caption = "Normality assessment for the average-value variables.")
```

```{r rq2-tests}
rq2_pearson <- cor.test(trxs$avg_val_received_log, trxs$avg_val_sent_log,
                        method = "pearson", alternative = "greater")
rq2_kendall <- cor.test(trxs$avg_val_received, trxs$avg_val_sent,
                        method = "kendall", alternative = "greater")
rq2_pearson
```

**Finding.** A significant positive correlation —
*r* = `r round(rq2_pearson$estimate, 3)`,
τ = `r round(rq2_kendall$estimate, 3)`
(both *p* `r format.pval(rq2_pearson$p.value, eps = .001)`) — addresses receiving
larger average values also send larger average values. Addresses do operate at a
characteristic value scale, which is exactly why the *conditional* questions of the
regression chapters (what happens at a *fixed* activity level) are the interesting
ones.

## RQ3 — Does sending cadence differ between fraudulent and legitimate addresses?

From value to time. Chapter 2's behavioural sections suggested the two populations
move to different rhythms; this is the formal version of that comparison. Only
addresses with at least one sent transaction — a defined cadence — enter it.

- **H0:** the average time between sent transactions does not differ between
  fraudulent and legitimate addresses.
- **Ha:** the average time between sent transactions differs between fraudulent and
  legitimate addresses.

```{r rq3-descriptives}
rq3 <- trxs %>% filter(is.finite(avg_min_sent_log), avg_min_between_sent_tnx > 0)

rq3 %>%
  group_by(flag_f) %>%
  summarise(n = n(),
            mean_log = mean(avg_min_sent_log),
            sd_log = sd(avg_min_sent_log),
            median_minutes = median(avg_min_between_sent_tnx)) %>%
  knitr::kable(digits = 2,
               caption = "Sending cadence by group.")
```

```{r rq3-tests}
rq3_levene <- leveneTest(avg_min_sent_log ~ flag_f, data = rq3)
rq3_welch  <- t.test(avg_min_sent_log ~ flag_f, data = rq3, var.equal = FALSE)
rq3_mwu    <- wilcox.test(avg_min_between_sent_tnx ~ flag_f, data = rq3)

# Effect sizes: Cohen's d (pooled) and rank-biserial r for the MWU check
g <- split(rq3$avg_min_sent_log, rq3$flag_f)
n1 <- length(g[[1]]); n2 <- length(g[[2]])
sp <- sqrt(((n1 - 1) * var(g[[1]]) + (n2 - 1) * var(g[[2]])) / (n1 + n2 - 2))
rq3_d  <- (mean(g[[1]]) - mean(g[[2]])) / sp
rq3_rb <- 1 - 2 * rq3_mwu$statistic / (n1 * n2)

rq3_levene
rq3_welch
```

**Finding.** Levene's test indicates unequal variances
(*p* `r format.pval(rq3_levene[1, "Pr(>F)"], eps = .001)`), so Welch's *t*-test is
the primary test: the cadence difference between groups is statistically
significant (*t* = `r round(rq3_welch$statistic, 2)`,
*p* `r format.pval(rq3_welch$p.value, eps = .001)`; Cohen's
*d* = `r round(rq3_d, 2)`), and the Mann–Whitney robustness check agrees
(rank-biserial *r* = `r round(as.numeric(rq3_rb), 2)`). Fraudulent and legitimate
addresses transact on measurably different rhythms — a genuinely *behavioural*
fraud signal, independent of value, and one that returns as a predictor in
Chapter 5.

## RQ4 — Does Ether outflow differ across dominant sent-token types?

From individual behaviour to ecosystems. If token communities differ in how much
Ether their addresses move, "which token dominates an address" is context that
later models should not ignore. Token types with fewer than 40 addresses are
excluded (unstable group estimates), as is `"None"` — no ERC20 activity is a
different phenomenon than *which* token dominates.

- **H0:** total Ether sent does not differ across the ERC20 token types most
  frequently sent from addresses.
- **Ha:** total Ether sent differs across the ERC20 token types most frequently
  sent from addresses.

```{r rq4-prep}
rq4 <- trxs %>%
  filter(erc20_most_sent_token_type != "None") %>%
  add_count(erc20_most_sent_token_type, name = "type_n") %>%
  filter(type_n >= 40) %>%
  mutate(token = factor(erc20_most_sent_token_type))

count(rq4, token, sort = TRUE) %>%
  knitr::kable(caption = "Token types entering the comparison (n ≥ 40).")
```

```{r rq4-box}
#| fig-cap: "Ether outflow by dominant sent-token type, on the log scale. Both the medians and the spreads differ across ecosystems — the spread differences are why a variance-robust test is used below."
ggplot(rq4, aes(x = reorder(token, total_ether_sent_log, median),
                y = total_ether_sent_log)) +
  geom_boxplot(fill = "grey85", outlier.alpha = 0.25) +
  coord_flip() +
  labs(title = "Different token ecosystems, different outflow profiles",
       x = NULL, y = "log10(total Ether sent + 1)") +
  theme_minimal()
```

```{r rq4-tests}
rq4_levene  <- leveneTest(total_ether_sent_log ~ token, data = rq4)
rq4_welch   <- oneway.test(total_ether_sent_log ~ token, data = rq4,
                           var.equal = FALSE)          # Welch's ANOVA
rq4_kruskal <- kruskal.test(total_ether_sent_log ~ token, data = rq4)
rq4_welch
```

```{r rq4-posthoc}
# Games–Howell post-hoc (appropriate under unequal variances)
if (requireNamespace("rstatix", quietly = TRUE)) {
  rstatix::games_howell_test(rq4, total_ether_sent_log ~ token) %>%
    filter(p.adj < 0.05) %>%
    arrange(p.adj) %>%
    head(10) %>%
    knitr::kable(digits = 3,
                 caption = "Significant Games–Howell pairwise contrasts (top 10).")
} else {
  cat("Install `rstatix` for the Games–Howell post-hoc table.")
}
```

**Finding.** With unequal variances
(Levene *p* `r format.pval(rq4_levene[1, "Pr(>F)"], eps = .001)`), Welch's ANOVA
finds a significant token-type effect on Ether outflow
(*F* = `r round(rq4_welch$statistic, 2)`,
*p* `r format.pval(rq4_welch$p.value, eps = .001)`), confirmed by Kruskal–Wallis
(*p* `r format.pval(rq4_kruskal$p.value, eps = .001)`); the post-hoc contrasts
identify which specific token ecosystems are associated with heavier outflows.

## RQ5 — Is the dominant received token associated with fraud?

The final question turns the token lens onto the label itself. Low-frequency token
types (< 40 addresses) are aggregated into `"Low Count"` so that expected cell
counts support the χ² approximation; `"None"` is retained as a meaningful category.

- **H0:** there is no association between the most frequently received ERC20 token
  type and the fraud flag.
- **Ha:** there is an association between the most frequently received ERC20 token
  type and the fraud flag.

```{r rq5-tests}
rq5 <- trxs %>%
  add_count(erc20_most_rec_token_type, name = "type_n") %>%
  mutate(token_rec = ifelse(type_n < 40, "Low Count", erc20_most_rec_token_type))

rq5_tab  <- table(rq5$token_rec, rq5$flag_f)
rq5_chi  <- chisq.test(rq5_tab)
rq5_cram <- sqrt(rq5_chi$statistic /
                 (sum(rq5_tab) * (min(dim(rq5_tab)) - 1)))   # Cramér's V
rq5_chi
```

```{r rq5-table}
prop.table(rq5_tab, margin = 1) %>%
  as.data.frame.matrix() %>%
  tibble::rownames_to_column("token") %>%
  arrange(desc(Fraudulent)) %>%
  head(10) %>%
  knitr::kable(digits = 3,
               caption = "Fraud share by dominant received token — the ten most fraud-associated types.")
```

**Finding.** The association is statistically significant
(χ² = `r round(rq5_chi$statistic, 1)`, df = `r rq5_chi$parameter`,
*p* `r format.pval(rq5_chi$p.value, eps = .001)`) with effect size Cramér's
*V* = `r round(as.numeric(rq5_cram), 2)`: *which* token dominates an address's
inflow is informative about its fraud status. The table above also foreshadows a
theme Chapter 5 examines critically — some token types are so strongly
fraud-associated that the question becomes *why*, and the answer turns out to
matter a great deal for any predictive use of this data.

## What the tests established

```{r summary-table, echo=FALSE}
tribble(
  ~RQ, ~Question, ~`Primary test`, ~`Robustness`, ~Verdict,
  "1", "Ether sent × transaction count", "Pearson (log)", "Kendall's τ", "Ha: positive, weak–moderate",
  "2", "Avg value received × sent", "Pearson (log, one-sided)", "Kendall's τ", "Ha: positive, strong",
  "3", "Sending cadence × fraud", "Welch's t (log)", "Mann–Whitney U", "Ha: groups differ",
  "4", "Ether sent × sent-token type", "Welch's ANOVA + Games–Howell", "Kruskal–Wallis", "Ha: token effect",
  "5", "Received-token type × fraud", "χ² independence + Cramér's V", "—", "Ha: associated"
) %>% knitr::kable(caption = "All five hypotheses in one view. In every case the robustness check agreed with the primary test.")
```

Five questions, five rejections of the null — and, more importantly for what
follows, none of them fragile: every verdict was confirmed by an estimator that
does not share the primary test's assumptions. Two threads now lead onward. The
behavioural and value findings (RQ1–3) set up Chapter 4's *conditional* question —
at a fixed activity level, what remains in an address's balance, and does fraud
shift it? And the token findings (RQ4–5) set up Chapter 5's — do the behavioural
signals predict the flag, and does token context change what they mean? Everything
here is inference about associations; prediction, with all its extra honesty
requirements, comes next.
