---
title: "5 · Predicting fraud"
subtitle: "From behavioural signals to flag probabilities — read with a fraud analyst's skepticism"
---

Chapter 3 established that *which* token dominates an address's inflow is
associated with its fraud status. This chapter turns from association to
conditional prediction: **can transaction behaviour — how much an address
transacts, how much of that is token activity, and how quickly it sends — predict
the likelihood of a fraud flag? And does the token ecosystem change what those
behaviours mean?** The goal is inferential understanding of the signals, not a
deployable classifier; that distinction ends up mattering more here than in any
other chapter.

## The analysis sample — and what the missing data quietly revealed

```{r setup}
library(dplyr)
library(ggplot2)
library(broom)
library(car)
library(pROC)

trxs <- readRDS("data/eth_clean.rds")

lr <- trxs %>%
  filter(!is.na(total_erc20_tnxs)) %>%
  add_count(erc20_most_rec_token_type, name = "type_n") %>%
  mutate(
    token_rec  = ifelse(type_n < 40, "Low Count", erc20_most_rec_token_type),
    token_rec  = relevel(factor(token_rec), ref = "None"),
    tnx_log    = log10(total_transactions + 1),
    erc20_log  = log10(total_erc20_tnxs + 1),
    avgmin_log = log10(avg_min_between_sent_tnx + 1)
  )

n_excl  <- sum(is.na(trxs$total_erc20_tnxs))
excl_fr <- sum(trxs$flag[is.na(trxs$total_erc20_tnxs)])
cat(nrow(lr), "addresses in sample;", sum(lr$flag), "flagged",
    sprintf("(%.1f%%; majority-class baseline accuracy %.1f%%)",
            100 * mean(lr$flag), 100 * (1 - mean(lr$flag))))
```

The model needs the ERC20 aggregates, so the `r n_excl` addresses lacking them are
excluded — and that routine step surfaced something worth pausing on. The fraud
share drops from 22.1% to `r sprintf("%.1f%%", 100 * mean(lr$flag))`, which is a
large move for a 8% exclusion, and the reason is stark: **all `r n_excl` of the
excluded addresses are flagged fraudulent** (`r excl_fr`/`r n_excl`). In other
words, in this dataset "ERC20 data missing" perfectly predicts the label. The most
plausible cause is mechanical — the fraud-address list appears to have been
assembled separately and joined in, with the ERC20 aggregates never computed for
part of that batch — which makes missingness an artefact of dataset construction,
not a behaviour of fraudsters. The practical consequences: missingness must never
be used as a feature on this data, and everything below speaks for token-active
addresses only. Low-frequency token types (< 40 addresses) are pooled into a
`"Low Count"` level so group estimates stay stable, and `"None"` — no dominant
ERC20 inflow — is kept as the meaningful reference category.

## Model 1: what behaviour alone says

```{r m1}
m1 <- glm(flag ~ tnx_log + erc20_log + avgmin_log,
          family = binomial, data = lr)

tidy(m1, exponentiate = TRUE, conf.int = TRUE) %>%
  knitr::kable(digits = 3,
               caption = "Model 1 — odds ratios with 95% CIs. Each OR is the multiplicative change in fraud odds for a ×10 change in the predictor.")
```

```{r m1-inline, include=FALSE}
or1 <- exp(coef(m1))
null_model <- glm(flag ~ 1, family = binomial, data = lr)
mcfadden <- function(m) 1 - as.numeric(logLik(m)) / as.numeric(logLik(null_model))
```

Read on the ×10 scale, the three effects sketch a coherent behavioural
fingerprint. A ×10 increase in **ordinary transactions** multiplies the fraud odds
by `r round(or1[["tnx_log"]], 2)` — sustained, plain on-chain activity is what
legitimacy looks like. A ×10 increase in **ERC20 token transactions** multiplies
the odds by `r round(or1[["erc20_log"]], 1)` — the flagged population is
token-transfer-heavy. And a ×10 *longer* **average gap between sent transactions**
multiplies the odds by `r round(or1[["avgmin_log"]], 2)`: quicker sending rhythms
lean fraudulent, echoing the cadence difference found in Chapter 3. Fraud, in this
data, looks token-heavy, plain-transaction-light, and fast. The model's McFadden
pseudo-R² is `r round(mcfadden(m1), 2)` and the VIFs below confirm the three
signals are not proxies for one another.

```{r m1-vif}
vif(m1) %>% knitr::kable(col.names = "VIF",
                         caption = "Variance inflation factors — the three behavioural signals carry distinct information.")
```

## From scores to decisions

A logistic model outputs probabilities; a fraud screen must output decisions, and
the gap between those two is where the operational honesty lives.

```{r m1-roc}
#| fig-cap: "Model 1 discrimination. AUC summarises ranking quality across all possible thresholds — it says nothing about any particular operating point."
roc1 <- roc(lr$flag, fitted(m1), quiet = TRUE)
plot(roc1, legacy.axes = TRUE, col = "#2C7FB8",
     main = sprintf("Behaviour-only model — AUC %.2f", auc(roc1)))
```

```{r m1-cm}
pred1 <- factor(ifelse(fitted(m1) >= 0.5, "Fraudulent", "Legitimate"),
                levels = c("Legitimate", "Fraudulent"))
cm <- table(pred1, lr$flag_f)

data.frame(
  ` ` = c("Predicted: Legitimate", "Predicted: Fraudulent"),
  `Actual: Legitimate` = cm[, "Legitimate"],
  `Actual: Fraudulent` = cm[, "Fraudulent"],
  check.names = FALSE
) %>%
  knitr::kable(row.names = FALSE,
               caption = "Confusion matrix at the default 0.5 threshold.")

tibble(
  metric = c("AUC", "Accuracy", "Sensitivity (fraud recall)", "Specificity"),
  value = c(as.numeric(auc(roc1)),
            sum(diag(cm)) / sum(cm),
            cm["Fraudulent", "Fraudulent"] / sum(cm[, "Fraudulent"]),
            cm["Legitimate", "Legitimate"] / sum(cm[, "Legitimate"]))
) %>% knitr::kable(digits = 3, caption = "Operating characteristics at the 0.5 threshold.")
```

The matrix tells the story the accuracy number hides. At the default threshold the
model catches `r cm["Fraudulent", "Fraudulent"]` of the
`r sum(cm[, "Fraudulent"])` flagged addresses
(sensitivity `r sprintf("%.0f%%", 100 * cm["Fraudulent","Fraudulent"] / sum(cm[,"Fraudulent"]))`)
while raising only `r cm["Fraudulent", "Legitimate"]` false alarms — and still
posts `r sprintf("%.0f%%", 100 * sum(diag(cm)) / sum(cm))` accuracy, because with
85% of addresses legitimate, accuracy mostly rewards saying "legitimate". The 0.5
threshold is a statistical default, not a policy: in a real fraud operation the
threshold is set by the relative cost of a missed fraud versus an investigation,
and lowering it here would trade a manageable rise in false alarms for a large
gain in recall. The behavioural signals rank addresses usefully
(AUC `r round(as.numeric(auc(roc1)), 2)`); what to *do* with the ranking is a
business decision.

## Model 2: does the token ecosystem change what behaviour means?

Chapter 3 found the dominant received token *associated* with fraud; the sharper
question is whether it *moderates* the behavioural signals — whether the same
pattern of activity means something different inside different token communities.
Model 2 interacts all three behavioural predictors with the token factor.

```{r m2}
m2 <- glm(flag ~ (tnx_log + erc20_log + avgmin_log) * token_rec,
          family = binomial, data = lr)

# Wald intervals: several token × behaviour cells contain (almost) only one
# class, so profile-likelihood CIs do not converge for those terms.
tidy(m2) %>%
  mutate(OR = exp(estimate),
         conf.low = exp(estimate - 1.96 * std.error),
         conf.high = exp(estimate + 1.96 * std.error)) %>%
  select(term, OR, conf.low, conf.high, p.value) %>%
  filter(p.value < 0.05) %>%
  arrange(p.value) %>%
  knitr::kable(digits = 3,
               caption = "Model 2 — significant terms only (Wald 95% CIs; full table in the folded code).")
```

The table already hints at what is happening: token-level odds ratios in the
hundreds of thousands, interaction ORs indistinguishable from zero. Estimates that
extreme are the signature of near-empty cells — token groups where one class
barely occurs — pushing coefficients toward the boundary. They are shown for
completeness, not interpreted individually; the trustworthy comparison is at the
model level.

```{r compare}
lrt <- anova(m1, m2, test = "Chisq")
roc2 <- roc(lr$flag, fitted(m2), quiet = TRUE)

tibble(model = c("M1: behaviour only", "M2: + token-type interactions"),
       pseudo_R2 = c(mcfadden(m1), mcfadden(m2)),
       AUC = c(as.numeric(auc(roc1)), as.numeric(auc(roc2))),
       AIC = c(AIC(m1), AIC(m2))) %>%
  knitr::kable(digits = 3,
               caption = "Model comparison. The likelihood-ratio test for the added terms is decisive (p < .001) — the interesting question is why.")
```

```{r roc-both}
#| fig-cap: "Both models' ROC curves. The jump is real in-sample — and the next paragraphs explain why it should not be taken at face value."
plot(roc1, col = "#2C7FB8", legacy.axes = TRUE,
     main = "Adding token context: dramatic — and suspicious")
plot(roc2, col = "#D95F02", add = TRUE)
legend("bottomright",
       legend = c(sprintf("M1: behaviour only (AUC %.2f)", auc(roc1)),
                  sprintf("M2: + token context (AUC %.2f)", auc(roc2))),
       col = c("#2C7FB8", "#D95F02"), lwd = 2, bty = "n")
```

```{r leakage, include=FALSE}
perfect <- lr %>%
  group_by(token_rec) %>%
  summarise(n = n(), fraud_rate = mean(flag), .groups = "drop") %>%
  filter(fraud_rate == 1)
```

### Reading the jump carefully

An AUC of `r round(as.numeric(auc(roc2)), 2)` from a logistic regression should
raise an eyebrow before it raises applause, and here the skepticism pays off. Part
of the improvement is genuine segment structure — Chapter 3's association result
made that expected. But a decisive part of it is simpler:
**`r nrow(perfect)` token levels in this sample coincide perfectly with the fraud
flag** — `r paste(sprintf("*%s* (n = %d)", perfect$token_rec, perfect$n), collapse = " and ")`
are flagged fraudulent without exception. Both are known spam-airdrop tokens, and
the likely mechanism runs through how the dataset was built: the fraud list appears
to draw heavily on addresses swept up in those scam ecosystems, so receiving these
tokens is partly a *consequence* of how the label was constructed rather than an
independent behavioural signal. Combined with the missingness artefact found
earlier, the message is consistent: this dataset rewards models that memorise its
construction.

For anyone taking this analysis toward a production model, three practices follow
naturally: validate out-of-time rather than in-sample, so construction artefacts
stop paying; ablate token *identity* and check how much discrimination survives on
behaviour alone (here, an AUC of `r round(as.numeric(auc(roc1)), 2)`); and treat
token context as segment information whose meaning must be re-learned as scam
ecosystems change, not as a stable feature. Read that way, Model 2 is best
understood as strong evidence that token context *carries* signal — and a worked
example of why a fraud analyst never deploys an in-sample 0.98.

## What this chapter established

Behaviour alone ranks addresses usefully — token-heavy, plain-transaction-light,
fast-cadence profiles carry materially higher flag odds — and token context adds
real but contaminated lift whose honest assessment requires validation this
cross-sectional dataset cannot provide. The estimates speak for token-active
addresses, against a label of undocumented provenance, in-sample only; the
deployment-grade treatment of this problem — held-out protocols, imbalance
handling, cost-sensitive thresholds — is the subject of the dedicated
machine-learning project that this suite's conclusions point toward.
